---
title: "Gerrymandering Analysis"
format: 
  dashboard:
    scrolling: true
execute:
  echo: true
  warning: false
  error: false
jupyter: r
---

### **2024 Election Results and the 2024 District Map**

```{r}
election_metrics <- merged |>
    filter(district_type == "CNG", votes > 0) |>
    group_by(CDDIST) |>
    summarize(
        total_district_votes = first(total_votes),
        dem_votes = sum(votes[grepl("dem", race_type)]),
        rep_votes = sum(votes[grepl("rep", race_type)]),
        dem_vote_share = dem_votes / (dem_votes + rep_votes),
        rep_vote_share = rep_votes / (dem_votes + rep_votes),
        dem_won = dem_votes > rep_votes
    ) |>
ungroup()

# MEAN MEDIAN
mean_dem_share <- mean(election_metrics$dem_vote_share)

# Median vote share
median_dem_share <- median(election_metrics$dem_vote_share)

# Mean-median score
mean_median_score <- mean_dem_share - median_dem_share

cat("Mean Democratic Vote Share:", round(mean_dem_share * 100, 2), "%\n")
cat("Median Democratic Vote Share:", round(median_dem_share * 100, 2), "%\n")
cat("Mean-Median Score:", round(mean_median_score * 100, 2), "%\n")
if (mean_median_score > 0) {
    cat("Interpretation: Small Democratic advantage (mean > median)\n\n")
} else {
    cat("Interpretation: Small Republican advantage (mean < median)\n\n")
}

# EFFICIENCY GAP
election_metrics <- election_metrics %>%
    mutate(
        vote_margin = abs(dem_votes - rep_votes),
        dem_wasted = if_else(
            dem_won,
            dem_votes - (dem_votes + rep_votes) / 2,
            dem_votes
        ),
        rep_wasted = if_else(
            !dem_won,
            rep_votes - (dem_votes + rep_votes) / 2,
            rep_votes
        )
    )

total_dem_wasted <- sum(election_metrics$dem_wasted)
total_rep_wasted <- sum(election_metrics$rep_wasted)
total_votes_cast <- sum(election_metrics$dem_votes + election_metrics$rep_votes)
efficiency_gap <- (total_rep_wasted - total_dem_wasted) / total_votes_cast

cat("EFFICIENCY GAP")
cat("Total Democratic Wasted Votes:", format(total_dem_wasted, big.mark = ","), "\n")
cat("Total Republican Wasted Votes:", format(total_rep_wasted, big.mark = ","), "\n")
cat("Efficiency Gap:", round(efficiency_gap * 100, 2), "%\n")
if (abs(efficiency_gap) < 0.08) {
    cat("Interpretation: Within acceptable range (< 8%)\n")
} else if (efficiency_gap > 0) {
    cat("Interpretation: Republicans have an efficiency advantage (wasting fewer votes)\n")
} else {
    cat("Interpretation: Democrats have an efficiency advantage (wasting fewer votes)\n")
}

cat("\nWon:\n")
cat("Democrats:", sum(election_metrics$dem_won), "seats\n")
cat("Republicans:", sum(!election_metrics$dem_won), "seats\n")

district24 <- st_read("gerrymandering-YaBoiPasta27/data/shapefiles/tl_2024_06_cd119.shp")
district24 <- select(district24, 'CD119FP', 'geometry')
election_2024 <- precinct_clean %>%
    mutate(party = if_else(grepl("dem", race_type), "house_dem", "house_rep")) %>%
    group_by(CDDIST, party) %>%
    summarize(votes = sum(votes), .groups = "drop") %>%
    pivot_wider(names_from = party, values_from = votes, values_fill = 0) %>%
    mutate(
        dem_votes = house_dem,
        rep_votes = house_rep,
        total_votes = dem_votes + rep_votes,
        dem_vote_share = dem_votes / total_votes,
        dem_won = dem_votes > rep_votes,
        CDDIST = sprintf("%02d", CDDIST)
    )

# Join with shapefile
districts_2024 <- district24 %>%
    left_join(election_2024, by = c("CD119FP" = "CDDIST")) %>%
    mutate(winner = if_else(dem_won, "Democrat", "Republican", missing = "Unknown"))

# Now you can map it
ggplot(districts_2024) +
    geom_sf(aes(fill = winner), color = "white", size = 0.2) +
    scale_fill_manual(
        values = c("Democrat" = "#2E74C0", "Republican" = "#CB454A"),
        na.value = "gray80"
    ) +
    theme_void() +
    labs(title = "2024 Congressional Districts", fill = NULL)
```

### **2025 Theoretical Election Results and the Proposed 2025 District Map**

```{r}
election_metrics <- district_votes %>%
    pivot_wider(
        names_from = party,
        values_from = total_votes,
        values_fill = 0
    ) %>%
    mutate(
        dem_votes = house_dem,
        rep_votes = house_rep,
        total_district_votes = dem_votes + rep_votes,
        dem_vote_share = dem_votes / (dem_votes + rep_votes),
        rep_vote_share = rep_votes / (dem_votes + rep_votes),
        dem_won = dem_votes >= rep_votes
    )

# MEAN MEDIAN
mean_dem_share <- mean(election_metrics$dem_vote_share, na.rm = TRUE)

# Median vote share
median_dem_share <- median(election_metrics$dem_vote_share, na.rm = TRUE)

# Mean-median score
mean_median_score <- mean_dem_share - median_dem_share

cat("MEAN-MEDIAN ANALYSIS\n")
cat("Mean Democratic Vote Share:", round(mean_dem_share * 100, 2), "%\n")
cat("Median Democratic Vote Share:", round(median_dem_share * 100, 2), "%\n")
cat("Mean-Median Score:", round(mean_median_score * 100, 2), "%\n")

# EFFICIENCY GAP
election_metrics <- election_metrics %>%
    mutate(
        vote_margin = abs(dem_votes - rep_votes),
        dem_wasted = if_else(
            dem_won,
            dem_votes - (dem_votes + rep_votes) / 2,
            dem_votes
        ),
        rep_wasted = if_else(
            !dem_won,
            rep_votes - (dem_votes + rep_votes) / 2,
            rep_votes
        )
    )

total_dem_wasted <- sum(election_metrics$dem_wasted, na.rm = TRUE)
total_rep_wasted <- sum(election_metrics$rep_wasted, na.rm = TRUE)
total_votes_cast <- sum(election_metrics$dem_votes + election_metrics$rep_votes, na.rm = TRUE)
efficiency_gap <- (total_rep_wasted - total_dem_wasted) / total_votes_cast

cat("EFFICIENCY GAP\n")
cat("Total Democratic Wasted Votes:", format(round(total_dem_wasted), big.mark = ","), "\n")
cat("Total Republican Wasted Votes:", format(round(total_rep_wasted), big.mark = ","), "\n")
cat("Efficiency Gap:", round(efficiency_gap * 100, 2), "%\n")

cat("SEATS WON\n")
cat("Democrats:", sum(election_metrics$dem_won, na.rm = TRUE), "seats\n")
cat("Republicans:", sum(!election_metrics$dem_won, na.rm = TRUE), "seats\n")
cat("Total Districts:", nrow(election_metrics), "\n")

district_votes_wide <- district_vote %>%
    filter(!is.na(party)) %>%
    mutate(DISTRICT = sprintf("%02d", as.numeric(DISTRICT))) %>%
    pivot_wider(
        names_from = party,
        values_from = total_votes,
        values_fill = 0
    ) %>%
    mutate(
        dem_votes = house_dem,
        rep_votes = house_rep,
        total_votes = dem_votes + rep_votes,
        dem_vote_share = dem_votes / total_votes,
        dem_won = dem_votes > rep_votes
    )

# Join with shapefile
districts_new_combined <- newdistricts %>%
    left_join(district_votes_wide, by = c("DISTRICT" = "DISTRICT")) %>%
    mutate(winner = if_else(dem_won, "Democrat", "Republican", missing = "Unknown"))

# Now you can map it
ggplot(districts_new_combined) +
    geom_sf(aes(fill = winner), color = "white", size = 0.2) +
    scale_fill_manual(values = c("Democrat" = "#2E74C0", "Republican" = "#CB454A"),
                      na.value = "gray80") +
    theme_void() +
    labs(title = "Proposed Districts", fill = NULL)
```

### **Maps**

Methodology

Data Sources: 2024 votes from CA Statewide Database; 2024 districts from Census TIGER; proposed districts from AB 604.

Method: Area-weighted interpolation allocated precinct votes to new districts proportionally by geometric overlap, assuming uniform voter distribution.

Metrics: Mean-median measures vote packing (±2% neutral); Efficiency gap measures wasted votes (±8% threshold).

Caveats: Single metrics don't prove gerrymandering; geographic clustering affects results; projections assume stable voter preferences.
