---
title: Exploratory Data Analysis
format: typst
---

### Question 1
Closest House race?

### Answer 1

```{r}
closest_house_races <- merged |>
    filter(district_type == "CNG", votes > 0) |>
    group_by(CDDIST) |>
    summarize(
        winner = name[which.max(votes)],
        winner_votes = max(votes),
        runner_up_votes = sort(votes, decreasing = TRUE)[2],
        margin = max(votes) - sort(votes, decreasing = TRUE)[2],
        margin_pct = (margin / total_votes[1]) * 100
    ) |>
    arrange(margin)
closest_house_races
```

closest race was CA45, the winner being Derek Tran, who won by 694 votes

### Question 2

How did incumbents perform compared to non-incumbents in House races?

### Answer 2

```{r}
incumbent_performance <- merged |>
    filter(district_type == "CNG", votes > 0) |>
    group_by(CDDIST) |>
    mutate(
        vote_share = votes / sum(votes) * 100,
        won = votes == max(votes)
    ) |>
    ungroup() |>
    group_by(incumbent) |>
    summarize(
        n_candidates = n(),
        avg_vote_share = mean(vote_share),
        win_rate = sum(won) / n() * 100,
        total_votes = sum(votes)
    )
incumbent_performance
```

incumbents had a huge advantage, winning 93.5% of their reelection bids
### Question 3

Which districts showed the strongest partisan lean (largest gap between Democratic and Republican performance)?

### Answer 3

```{r}
partisan_lean <- merged |>
    filter(district_type %in% c("PRES", "CNG", "SEN"), votes > 0) |>
    mutate(party = ifelse(grepl("dem", race_type), "DEM", "REP")) |>
    group_by(CDDIST, party) |>
    summarize(
        total_party_votes = sum(votes),
        .groups = "drop"
    ) |>
    pivot_wider(
        names_from = party,
        values_from = total_party_votes,
        values_fill = 0
    ) |>
    mutate(
        total_votes = DEM + REP,
        dem_pct = (DEM / total_votes) * 100,
        rep_pct = (REP / total_votes) * 100,
        partisan_lean = dem_pct - rep_pct,
        lean_category = case_when(
        partisan_lean > 20 ~ "Strongly Democratic",
        partisan_lean > 10 ~ "Lean Democratic",
        partisan_lean > -10 ~ "Competitive",
        partisan_lean > -20 ~ "Lean Republican",
        TRUE ~ "Strongly Republican"
    )
  ) |>
arrange(desc(abs(partisan_lean)))
partisan_lean
```
