---
title: "Practice Worksheet: Gerrymandering"
format: typst
---

Recall the 3 quantitative approaches to measuring gerrymandering discussed in lecture:\
1. Mean - Median Score (Grofman, 1983)\
2. Efficiency Gap (Stephanopolous and McGee, 2015)\
3. Ensembles / Extreme Outliers\
Use the following table to practice applying the first two techniques.\

| District | D Votes | R Votes | Result  |
|:---------:|:-------:|:-------:|:--------|
| 1 | 62 | 38 | D wins |
| 2 | 55 | 45 | D wins |
| 3 | 44 | 56 | R wins |
| 4 | 47 | 53 | R wins |
| 5 | 46 | 54 | R wins |
| **Total** | **254** | **246** |          |

```{r}
#| message: false
#| echo: false
library(tidyverse)
votes <- tibble(District = 1:5,
                d_votes = c(62, 55, 44, 47, 46),
                r_votes = c(38, 45, 56, 53, 54),
                result = c(rep("D wins", 2), rep("R wins", 3)))
```

### Q1. Mean-Median Score

Calculate the mean-median score associated with this particular district breakdown of votes.

```{r}
mean_median_score <- votes %>%
  mutate(d_share = d_votes / (d_votes + r_votes)) %>%   # Democratic vote share
  summarise(
    mean_share = mean(d_share),
    median_share = median(d_share),
    mean_median = mean_share - median_share
  )

mean_median_score
```

### Q2. Efficiency Gap

Calculate the Efficiency Gap (EG) associated with this particular district breakdown of votes.

*Tip*: it may be helpful to write a function that takes a vector of votes across districts for party A and vector of votes for the party B and returns a vector of the number votes wasted by party A across districts.


```{r}
wasted_votes <- function(party_votes, opponent_votes) {
  total_votes <- party_votes + opponent_votes
  ifelse(
    party_votes > opponent_votes,
    party_votes - floor(total_votes / 2) - 1,
    party_votes
  )
}

# Compute wasted votes for D and R
votes <- votes %>%
  mutate(
    d_wasted = wasted_votes(d_votes, r_votes),
    r_wasted = wasted_votes(r_votes, d_votes)
  )

# Compute efficiency gap
EG <- sum(votes$d_wasted - votes$r_wasted) / sum(votes$d_votes + votes$r_votes)
EG
```
